Crystal as a thermodynamic computer

### 2024-10-13 paper/main.tex
update
update one figure
Wrote down the problem, temporarily blocked part: Numerical simulation

In the rest of this paper, we first show the design of crystalline structure that can be programmed on its surfaces to perform universal computation in \Cref{sec:Overview-of-main-results}.
The method we use to study the thermalization process is introduced in \Cref{sec:model-method}.
Finally, we reported numerical simulation results in \Cref{sec:numerical-result}. We show that even when the encoded problem is a deterministic computation, the system is not guaranteed to thermalize in time polynomial to the size of the crystal.
However, if we use the moving temperature gradient method, the system can thermalize in time almost linear to the size of the crystal, i.e. just like a normal computer.

@@

\draw    (305.63,73.33) -- (329.63,73.33) ;
\draw [shift={(331.63,73.33)}, rotate = 180] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw   (283.2,286.4) .. controls (283.2,291.07) and (285.53,293.4) .. (290.2,293.4) -- (314.2,293.4) .. controls (320.87,293.4) and (324.2,295.73) .. (324.2,300.4) .. controls (324.2,295.73) and (327.53,293.4) .. (334.2,293.4)(331.2,293.4) -- (358.2,293.4) .. controls (362.87,293.4) and (365.2,291.07) .. (365.2,286.4) ;

@@

    \end{tikzpicture}
    \caption{A surface programmable thermodynamic computer (TC) that implements the rule 110 elementary cellular automata. The vertical direction is the space axis,
    while the horizontal direction is the time axis.
    The computation is done by moving the heat zone (orange colored area) from the input surface (left) to the output surface (right) slowly with a constant speed $v$.
    The TC has a translational invariant structure, and each unit cell (inset) is a $110$ rule cellular automata gadget that consists of 5 mutually correlated spins (white circles).
    % \jinguo{please add symbols for space, time and width}
    % \ym{I add symbols. please check if i still need to put original captions here}
    \jinguo{please make the circles and texts in the inset larger}
    }

@@

We can define an effective heat zone with width $W'$ as the width of the raising edge of the temperature distribution, where the spins can be cooled actively.
%Defining $\Delta E_{max}$ to be the largest possible energy change when a spin is fliped.
%The effective heat zone with width $W'(\Lambda)$ should then be $[y-W'(\Lambda), y]$, where position $y$ satisfying 
%$\exp(\frac{\Delta E_{max}}{T(y)}) = 0.1$ and $y<L(t)$. 
% Since the layers in fronts of the effective heat zone can only be fliped with a $\epsilon$ probability if the local energy would increase. 
The spins outside the effective heat zone are either frozen (too cold) or flip randomly (too hot).
As a spin flips with a probability $\sim e^{-\Delta E/T}$, the cold edge of the heat zone can be defined as the position where the probability of flipping is $\epsilon$.
We have $W'$ scales with the required error probability $\epsilon$ as:
\begin{equation}
    e^{-\frac{\Delta E}{T_{\rm max}}\Lambda^{W'}} \sim \epsilon
\end{equation}
\ym{But I still don't think the constant $c$ is reasonable in this version. Anyway it is a better 
description for $W'$.}
\jinguo{I changed $c$ to $1/c$}

@@

Let $f(x) = c x + b$, then $e^{f(W')} \sim \ln(\epsilon^{-1})^{\frac{c}{\ln(\Lambda)}}$.
Combining with \Cref{eq:totaltime} and the requirement of $\epsilon \sim \frac{W'}{M}$, the total annealing time can be estimated as
\begin{equation}
    \label{eq:total-annealing-time}
    t_{\rm total} \sim \frac{M}{W'}\ln\left(\frac{M}{W'}\right)^{1+ \frac{c}{\ln(\Lambda)}}
\end{equation}
The reason why we choose $f$ to be linear is due to the simplicity of the model, and it turns out to be a good approximation in the numerical simulation.
%To determine the relationship between the total annealing time and the input size $N$, we should
%further discuss the relationship between the spectral gap given in  \Cref{eq:spectral-gap} and inputs size $N$.
%However, we failed to find a direct relationship.
%Numerical result shows that computing complexity is linear to $N$.
%Thus a total sweep time for the computation should scale as $NM\log^{1 + \frac{c}{\ln(\Lambda)}}(M)$.
\ym{Please help me to think about how to evaluate the time complexity while $N$ increases.}\jinguo{Leave it to Maddie}
The computing speed $v$ defined as operations per unit time, can be estimated as
\begin{equation}
    \label{eq:moving-speed}
    v = \frac{M}{t_{total}} \sim \ln(M)^{-(1 + \frac{c}{\ln(\Lambda)})}
\end{equation}
As the computational step $M$ increases, the computing speed $v$ decreases logarithmically.

\section{Numerical simulation}\label{sec:numerical-result}
%Given that simulating a CA with a deep computational layer on the Ising model requires many spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.
Instead of studying the spin-glass model directly, we propose a toy energy model to enable large-scale numerical simulations.
%on a 2D lattice of size $N \times M$.
The energy model on a 2D lattice of size $N \times M$ is defined as:
\begin{equation}\label{eq:toy-hamiltonian}
E(\vn) = \sum_{j=1}^{M-1}\sum_{i=1}^{N} (n_{i+1, j} \land \neg(n_{i, j} \land \neg n_{i-1, j})) \veebar n_{i,j+1},
\end{equation}
where $n_{i, j} \in \{0, 1\}$ is the state of the cell at row $i$ and column $j$, the lattice is periodic in the space direction, i.e. $n_{0, t} = n_{N, j}$.
%$S$ to be the length of the material along the ``Space'' axis, $D$
%to be the length of the material along the ``Time'' axis. 
%Its ground state with energy $0$ also encodes a $110$ CA, where the direction of computation is from the left to right.

@@

>>>>

In the rest of this paper, we first show the design of crystalline structure that can be programmed on its surfaces to perform universal computation in \Cref{sec:Overview-of-main-results}.
The method we use to study the thermalization process is introduced in \Cref{sec:model-method}.
Finally, we reported numerical simulation results in \Cref{sec:numerical-result}. We show that even when the encoded problem is a deterministic computation, the system is not guaranteed to thermalize in time polynomial to the size of the crystal.
However, if we use the moving temperature gradient method, the system can thermalize in time almost linear to the size of the crystal, i.e. as efficient as a normal computer.

@@

% \draw    (305.63,73.33) -- (329.63,73.33) ;
% \draw [shift={(331.63,73.33)}, rotate = 180] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

% \draw   (283.2,286.4) .. controls (283.2,291.07) and (285.53,293.4) .. (290.2,293.4) -- (314.2,293.4) .. controls (320.87,293.4) and (324.2,295.73) .. (324.2,300.4) .. controls (324.2,295.73) and (327.53,293.4) .. (334.2,293.4)(331.2,293.4) -- (358.2,293.4) .. controls (362.87,293.4) and (365.2,291.07) .. (365.2,286.4) ;

@@

    \end{tikzpicture}
    \caption{A surface programmable thermodynamic computer (TC) that implements the rule 110 elementary cellular automata. The vertical direction is the space axis,
    while the horizontal direction is the time axis.
    The computation is done by moving the heat zone (orange colored area) from the input surface (left) to the output surface (right) slowly with a constant speed $v$.
    The TC has a translational invariant structure, and each unit cell (inset) is a $110$ rule cellular automata gadget that consists of 5 mutually correlated spins (white circles).
    % \jinguo{please add symbols for space, time and width}
    % \ym{I add symbols. please check if i still need to put original captions here}
    % \jinguo{please make the circles and texts in the inset larger}
    }

@@

We can define an effective heat zone with width $W'$ as the width of the raising edge of the temperature distribution, where the spins can be cooled actively.
%Defining $\Delta E_{max}$ to be the largest possible energy change when a spin is fliped.
%The effective heat zone with width $W'(\Lambda)$ should then be $[y-W'(\Lambda), y]$, where position $y$ satisfying 
%$\exp(\frac{\Delta E_{max}}{T(y)}) = 0.1$ and $y<L(t)$. 
% Since the layers in fronts of the effective heat zone can only be fliped with a $\epsilon$ probability if the local energy would increase. 
The spins outside the effective heat zone are either frozen (too cold) or flip randomly (too hot).
As a spin flips with a probability $\sim e^{-\Delta E/T}$, the cold edge of the heat zone can be defined as the position where the probability of flipping is $\epsilon$.
We have $W'$ scales with the required error probability $\epsilon$ as:
\begin{equation}
    e^{-\frac{\Delta E}{T_{\rm max}}\Lambda^{W'}} \sim \epsilon
\end{equation}
% \ym{But I still don't think the constant $c$ is reasonable in this version. Anyway it is a better 
% description for $W'$.}
% \jinguo{I changed $c$ to $1/c$}

@@

Let $f(x) = c x + b$, then $e^{f(W')} \sim \ln(\epsilon^{-1})^{\frac{c}{\ln(\Lambda)}}$.
Combining with \Cref{eq:totaltime} and the requirement of $\epsilon \sim \frac{W'}{M}$, the total annealing time can be estimated as
\begin{equation}
    \label{eq:total-annealing-time}
    t_{\rm total} \sim \frac{M}{W'}\ln\left(\frac{M}{W'}\right)^{1+ \frac{c}{\ln(\Lambda)}}
\end{equation}
The reason why we choose $f$ to be linear is due to the simplicity of the model, and it turns out to be a good approximation in the numerical simulation.
%To determine the relationship between the total annealing time and the input size $N$, we should
%further discuss the relationship between the spectral gap given in  \Cref{eq:spectral-gap} and inputs size $N$.
%However, we failed to find a direct relationship.
%Numerical result shows that computing complexity is linear to $N$.
%Thus a total sweep time for the computation should scale as $NM\log^{1 + \frac{c}{\ln(\Lambda)}}(M)$.
% \ym{Please help me to think about how to evaluate the time complexity while $N$ increases.}\jinguo{Leave it to Maddie}
The computing speed $v$ defined as operations per unit time, can be estimated as
\begin{equation}
    \label{eq:moving-speed}
    v = \frac{M}{t_{total}} \sim \ln(M)^{-(1 + \frac{c}{\ln(\Lambda)})}
\end{equation}
As the computational step $M$ increases, the computing speed $v$ decreases logarithmically.

\subsubsection{Question about the above theory}

\ym{Please check this part}

We expect the error rate would decrease exponentially with the annealing time while $N, M$ is fixed, i.e. 
\begin{equation}
    \ln(\frac{1}{\epsilon}) \propto t_{\rm total}
\end{equation}

However, we found that the error rate is only inversely proportional to the annealing time, i.e,
$\frac{1}{\epsilon} \propto t_{\rm total}$.

\begin{figure}[h]
    \centering
    \includegraphics[width=\columnwidth,keepaspectratio]{../notes/images/eps_vs_t.pdf}
    \caption{(a): The error rate $\epsilon$ scales inversely with the total annealing time $t_{\rm total}$, which is different from the expected exponential scaling.
    (b): The log plot implies that $\ln(v)$ is linear to $\ln(M)$, which then implies $v \propto (\frac{1}{M})^{b}$, where $b$
    is the slope of the line.}
    \label{questionplot}
\end{figure}

\Cref{questionplot}(b) shows that the moving speed $v$ is at least proportional to $\frac{1}{M}$,
 which means we need at least $O(M^2)$ time to achieve a fixed error rate. This is correspond to the relationship
 between $\epsilon$ and $t_{\rm total}$ when $N, M$ is fixed.

The main problem is that $\epsilon$ does not obey the hypothesis that it decreases exponentially with the annealing time.
We come up with this hypothesis based on the intuition that the sampling procedure is a Markov chain, 
the proportion of the non-steady-state distribution should decrease exponentially with the number of samples.

\ym{I think this hypothesis make sense, but how to check whether the code is correct or not?
I set a minimum temperature of $10^{-5}$ in simulation. Should this be the reason?}

This problem has nothing to do with the temperature distribution model. Same relationship between
$v$ and $M$ also exist when we choose to anneal this system with a gaussian type temperature distribution.

\iffalse
\section{Numerical simulation}\label{sec:numerical-result}
%Given that simulating a CA with a deep computational layer on the Ising model requires many spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.
Instead of studying the spin-glass model directly, we propose a toy energy model to enable large-scale numerical simulations.
%on a 2D lattice of size $N \times M$.
The energy model on a 2D lattice of size $N \times M$ is defined as:
\begin{equation}\label{eq:toy-hamiltonian}
E(\vn) = \sum_{j=1}^{M-1}\sum_{i=1}^{N} (n_{i+1, j} \land \neg(n_{i, j} \land \neg n_{i-1, j})) \veebar n_{i,j+1},
\end{equation}
where $n_{i, j} \in \{0, 1\}$ is the state of the cell at row $i$ and column $j$, the lattice is periodic in the space direction, i.e. $n_{0, t} = n_{N, j}$.
%$S$ to be the length of the material along the ``Space'' axis, $D$
%to be the length of the material along the ``Time'' axis. 
%Its ground state with energy $0$ also encodes a $110$ CA, where the direction of computation is from the left to right.

@@

\fi
